---
title: "Getting Started with civic.icarm and DataCitizen-Pro"
author: "Olushina Olawale Awe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with civic.icarm and DataCitizen-Pro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 120
)
```

## Introduction

The `civic.icarm` package implements the **ICARM** framework:
**I**nterpretable, **C**ivic-**A**ccountable, and **R**esponsible **M**achine Learning.

It is the computational backbone of the **DataCitizen-Pro** research programme —
a DFG-funded project at Ludwigsburg University of Education (LUE) developing
*data literacy*, *statistical reasoning*, and *democratic judgment* competencies
in civic and political teacher education.

This vignette walks through the complete workflow:

1. Split data reproducibly  
2. Train an interpretable model  
3. Explain the model globally  
4. Audit fairness across demographic groups  
5. Generate a civic accountability scorecard  

---

## Installation

```{r install, eval=FALSE}
# From GitHub (development)
remotes::install_github("oluwale-awe/civic.icarm")

# With all suggested packages for full functionality
remotes::install_github("oluwale-awe/civic.icarm", dependencies = TRUE)
```

```{r load}
library(civic.icarm)
library(ggplot2)
library(dplyr)
```

---

## 1. The Data: Civic Voting Participation

`civic_voting` is a synthetic dataset of 1,000 individuals, calibrated to
approximate German civic participation surveys. It includes both substantive
predictors (age, education, political interest) and protected attributes
(gender, migration background) for fairness analysis.

```{r data}
data(civic_voting)
glimpse(civic_voting)
```

```{r data-summary}
civic_voting |>
  count(voted, gender) |>
  tidyr::pivot_wider(names_from = voted, values_from = n)
```

**DataCitizen-Pro teaching point:** Before modelling, examine the raw group
disparities. Are there differences in voting rates by gender or migration
background? Should a model *replicate* or *correct* these patterns?

---

## 2. Reproducible Train/Test Split

```{r split}
splits <- civic_split(
  civic_voting,
  prop      = 0.75,
  seed      = 2025,
  stratify  = "voted"       # preserve class balance
)

cat("Training set:", nrow(splits$train), "rows\n")
cat("Test set    :", nrow(splits$test),  "rows\n")

# Verify stratification
prop.table(table(splits$train$voted))
prop.table(table(splits$test$voted))
```

---

## 3. Train an Interpretable Classifier (CART)

```{r model}
m_cart <- civic_fit_classification(
  formula = voted ~ age + education + political_interest +
                    news_consumption + civic_org_member + income,
  data    = splits$train,
  model   = "cart",
  seed    = 2025
)

print(m_cart)
```

The `civic_model` object stores the seed and a SHA-256 hash of the training
data — sufficient to reproduce the model exactly.

```{r predict}
# Predictions on test set
y_hat  <- predict(m_cart, newdata = splits$test, type = "class")
y_prob <- predict(m_cart, newdata = splits$test, type = "prob")[, "yes"]

# Evaluate
metrics <- civic_metrics(
  y_true  = splits$test$voted,
  y_pred  = y_hat,
  y_prob  = y_prob,
  positive = "yes"
)
round(metrics, 3)
```

---

## 4. Global Explanation: Feature Importance

```{r explain}
explainer <- civic_explain(m_cart, data = splits$train)
print(explainer)
```

```{r plot-importance, fig.cap="Feature importance from the CART classifier."}
civic_plot_importance(explainer) +
  labs(title = "What drives civic voting participation? (CART)",
       subtitle = "DataCitizen-Pro — civic.icarm")
```

**DataCitizen-Pro teaching point:** Political interest and education dominate.
Ask learners: *Is it appropriate to use education as a feature in a system
that allocates civic resources?*

---

## 5. Threshold Analysis

Classification threshold choice is a *democratic decision*, not merely a
technical one. The 0.5 default treats false positives and false negatives as
equally costly — but is that appropriate for civic applications?

```{r thresholds}
thr_tbl <- civic_thresholds(
  y_true    = splits$test$voted,
  y_prob    = y_prob,
  positive  = "yes"
)
```

```{r plot-thresholds, fig.cap="Performance metrics across classification thresholds."}
civic_plot_thresholds(thr_tbl,
  metrics = c("accuracy", "recall", "precision", "f1")
)
```

---

## 6. Fairness Audit

```{r fairness}
fair_gender <- civic_fairness_report(
  object    = m_cart,
  data      = splits$test,
  outcome   = "voted",
  protected = "gender",
  positive  = "yes"
)
print(fair_gender)
```

```{r plot-fairness-tpr, fig.cap="Equal opportunity: True positive rate by gender."}
civic_plot_fairness(
  fair_gender,
  metric   = "tpr",
  title    = "Equal Opportunity by Gender",
  ref_line = NULL
)
```

```{r plot-fairness-dp, fig.cap="Disparate impact: positive prediction rate by gender. Dashed line = 80% rule."}
civic_plot_fairness(
  fair_gender,
  metric   = "dp_ratio",
  title    = "Disparate Impact Ratio by Gender",
  ref_line = 0.8
)
```

**DataCitizen-Pro teaching point:** The EU AI Act (2024) mandates fairness
reporting for high-risk AI systems. Learners discuss: *Which metric — equal
opportunity, demographic parity, or equalized odds — is most appropriate for
a model advising on civic resource allocation?*

---

## 7. Confusion Matrix

```{r confusion, fig.cap="Confusion matrix on the test set."}
civic_plot_confusion(
  y_true = splits$test$voted,
  y_pred = y_hat
)
```

---

## 8. Audit Trail

```{r audit}
trail <- civic_audit_trail(
  object   = m_cart,
  metrics  = metrics,
  fairness = fair_gender,
  analyst  = "O. O. Awe",
  notes    = "Baseline CART model — DataCitizen-Pro pilot vignette"
)
cat(trail)
```

The JSON audit trail is a *machine-readable civic accountability document*.
In a real deployment, it would be published alongside the model.

---

## 9. Civic Scorecard (DataCitizen-Pro Synthesis)

```{r scorecard}
civic_scorecard(
  object    = m_cart,
  test_data = splits$test,
  outcome   = "voted",
  protected = "gender",
  positive  = "yes",
  analyst   = "O. O. Awe",
  project   = "DataCitizen-Pro Pilot 2025"
)
```

---

## 10. Comparing Models

```{r compare-logistic}
m_logit <- civic_fit_classification(
  voted ~ age + education + political_interest + news_consumption,
  data  = splits$train,
  model = "logistic",
  seed  = 2025
)

ex_logit <- civic_explain(m_logit)
civic_plot_importance(ex_logit,
  title = "Feature importance — Logistic Regression (|coefficient|)"
)
```

**DataCitizen-Pro discussion:** CART vs. logistic regression have different
interpretability affordances. CART produces a *rule set*; logistic regression
produces a *weighted score*. Which is more appropriate for a civic context?

---

## Session info

```{r session}
sessionInfo()
```
